---
title: 'Challenges in microbiome translational research'
subtitle: 'BioData meetup 4 | April 2026'
author: "Vivien Pichon"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    theme: flatly
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

## Load packages
### Always useful
library(tidyverse)   # Data manipulation
library(reshape2)    # Data reshaping
library(knitr)       # Formatting for the tables
library(kableExtra)  # Formatting for the tables

### Differential expression
library(edgeR)

### Data visualization
library(viridis)     # Colourblind-friendly colour scale
library(RColorBrewer) # Colour palette
library(pheatmap)    # Heatmap package
library(ggtree)      # Build the phylogenetic trees
library(ggtreeExtra) # Extra functionalities for ggtree
library(ggnewscale)  # Allows to add multiple heatmaps around a single tree


```

# 1 - Differential abundance strains detection
## Import data

```{r}

## Import ASV table
asvtab_filtered <- read.csv( "data/asvmat_br.csv",
                             check.names = FALSE,
                             row.names = 1)

## Turn it into a matrix
asvmat_BR <- as.matrix(asvtab_filtered)


## Import phylogeny tab
phylo_filtered <- read.csv( "data/phylogeny_br.csv",
                            check.names = FALSE,
                            row.names = 1)

## Turn it into a matrix
phylo_BR <- as.matrix(phylo_filtered)


## Import sample metadata
meta_BR <- read.csv("data/metadata_br.csv",
                     check.names = FALSE,
                     row.names = 1)

## Import reference sequences
sequences <- read.csv("data/ref_sequences_br.csv",
                     check.names = FALSE)

```

## Shape data

```{r}

## Create DGEList objects
### Bintje
dge_br <- DGEList(counts = asvmat_BR,
                 genes = phylo_BR,
                 samples = meta_BR)

## Convert type of the "treatment" column to factor
dge_br$samples$treat <- factor(dge_br$samples$treat,
                             levels = c("C", "V"))

## TMM normalization (Trimmed Mean of the M-values)
dge_br <- edgeR::calcNormFactors(dge_br)

```

## Find differentially abundant taxa (edgeR)

```{r Bintje_rhizosphere_DA}

## Create Multidimensional Scaling plot
plotMDS(dge_br
        , labels=rownames(dge_br$samples)
        , col=as.numeric(dge_br$samples$treat)
        , cex=0.5)

## Reporting
cat( sprintf("The color represents the infection status (red is infected). /nMost of the variability is linked to the infection status." ))

## Design the model's formula for the Generalized Linear Model
design <- model.matrix(~ treat , data = dge_br$samples )

## Estimate the variability of counts of each ASV (necessary for negative binomial GLM)
glmest <- estimateGLMRobustDisp(dge_br, design = design)

## Fits negative binomial GLM. prior.count adds a pseudocount to stabilize estimates for low-occurring ASVs.
fit_glm <- glmFit(glmest, design = design, prior.count = 0.125)

## GLM Likelihood Ratio test, checks if the infection status has an effect on the counts of each ASV.
## Calculates AND ADJUSTS p-value.
lrt_treat<- glmLRT(fit_glm, coef = 2)

## Sorts ASV per significance
tt_treat <- as.data.frame(topTags(lrt_treat, n = Inf ))

## Export differentially expressed ASVs counts 
write.csv(tt_treat, file = "data/all_ASV.csv")

```

We can then plot the results as a volcano plot.

```{r}

## Summarize differential abundance (used for the volcano plot below)
treat_summary <- t(
    summary( de <- decideTests(lrt_treat)) )

## Create volcano plot
plotMD( lrt_treat, main = "Differential abundance in the rhizosphere", cex=0.5 )
abline( h=c(-1,1), col = "blue" )
text( x = 12, y = -4, labels = paste0("Up: ",treat_summary[3], 
                                      "\n NotSign: ", treat_summary[2], 
                                      "\n Down: ", treat_summary[1]) )

```

In the plot above, each point represents one ASV.
The x axis represents the average abundance (log scale), and the y axis represents the log-fold change between the infected and uninfected plants.
The red dots represent the ASV that became more abundance upon infection, the blue dots are the ASV that became less abundant, the black dots did not change significantly.


Now that we have the list of interesting ASVs, we can get their sequences and look for them in the isolated strains collection.

```{r}
## Get the list of all the ASVs
all_asv <- rownames(tt_treat)

## Get their DNA sequence
all_seq <- sequences |> 
  filter(asv %in% all_asv) |> 
  column_to_rownames("asv")
# => Show it

## Create a function to export the sequences as fasta
export_to_fasta <- function(df, output_file) {
  # Extract row names and sequences
  # We replace spaces with underscores because BLAST+ reads sequence IDs up to the first space
  strain_names <- gsub(" ", "_", rownames(df)) 
  
  # Assuming the sequence is in the first column
  sequences <- as.character(df[, 1]) 
  
  # Create the >Header lines
  headers <- paste0(">", strain_names)
  
  # Interleave headers and sequences (Row 1: Header, Row 2: Seq, Row 3: Header...)
  fasta_lines <- c(rbind(headers, sequences))
  
  # Write the lines to the specified file
  writeLines(fasta_lines, output_file)
  
  cat("Successfully exported", length(strain_names), "sequences to", output_file, "\n")
}

## Save the sequences of all ASV as fasta
export_to_fasta(all_seq, "data/all_ASV.fasta")

```


# 2 - Community shifts visualization

Some more data wrangling is necessary at this step, we will skip it to focus on the tree creation and commentary:

- Creation of a giant phylogenetic tree with all the ASVs (in Unix on HPC, aligned with MAFFT, tree created with FastTree)
- Aggregate the taxa at the **Order** phylogenetic level.
- Use de abundance to calculate the **total abundance of each Order**.
- Use the **list of differentially** abundant ASV to calculate two indicators:
  - A **local indicator**: Proportion of differentially abundant ASVs (relative to the **group's abundance**)
  - A **global indicator**: Average log-fold change of differentially abundant ASVs (relative to the **total abundance**)

## Import and filter objects

```{r import_objects}
## phylogenetic trees
tree <- readRDS("data/tree_objects/ASV_phylo.rds")

## Phylogenetic information
tree_info <- readRDS("data/tree_objects/tree_info.rds")

## List of orders belonging to each phylum (used for colors)
taxalist <- readRDS("data/tree_objects/taxa_list.rds")

## Abundance matrix (used for annotations)
abund_matrix <- readRDS("data/tree_objects/abund_matrix.rds")

## Local indicator (described above,used for more annotations)
DA_proportion_matrix <- readRDS("data/tree_objects/abund_ratio_matrix.rds")

## Global indicator (described above, used for even more annotations)
logfc_direction_matrix <- readRDS("data/tree_objects/logfc_matrix.rds")

```

## Create base tree

```{r create_base_tree}

## Make basic tree
p <- ggtree(tree, 
            ladderize = TRUE, 
            layout="fan",
            open.angle = 2.5) %<+% info +
  geom_tiplab(size = 2.5, 
              align = TRUE, 
              linesize = 0.5,
              linetype = "dotted",
              show.legend = FALSE)+
  labs(title = "Generation 2, rhizosphere")+
  theme(legend.title = element_text(size = 16),
        legend.text = element_text(size = 14))

## rotate tree and add colors
q <- p  |>  rotate(230) |> rotate(225)|> rotate(228)|> rotate(224)|> rotate(215)|> rotate(217)|> rotate(219)|> rotate(205)|> rotate(201) |> rotate(213)|> rotate(226)|> rotate(214) |> rotate(194) |> rotate(189) |> rotate(190) |> rotate(163) |> rotate(164) |>
  groupOTU(taxalist, "Phylum") +
  aes(color = Phylum) +
  scale_color_manual(values = c("deeppink2", "orangered", "orange", "olivedrab3", "limegreen","springgreen4", "lightseagreen", "steelblue2", "royalblue1", "grey60","purple1", "orchid2", "hotpink1"))

```

## Add informative matrices

```{r}

r <- gheatmap(q,
         abund_matrix,
         offset=0.88,
         width =0.15,
         color = "white",
         colnames_angle = 45,
         font.size = 2.5,
         colnames_offset_y = 0,
         colnames_offset_x = 0
         ) +
  scale_fill_gradientn(
    colors=c("#FFFFFF", "#FDE725FF", 
             "#94D840FF", "#3CBC75FF", "#1F968BFF", "#2D718EFF", 
             "#3F4788FF", "#440154FF"),
    limits = c(0, 6e6),
                      name = "TMM-normalized \nAbundance",
                      na.value = "grey"
                      ) +
  new_scale_fill()

## Add density heatmap
s <- gheatmap(r,
              group_abund_ratio,
              offset=0.66,
              width = 0.15,
              color = "white",
              colnames_angle = 45,
              font.size = 2.5,
              colnames_offset_y = 0,
              colnames_offset_x = 0
              )+

  scale_fill_gradientn(colors = c("white", "#3F007D"),
                      limits = c(0, 1),
                      name = "Ratio",
                      na.value = "white"
                      ) +
  new_scale_fill()

## Add averaged logfc heatmap
t <- gheatmap(s,
         logfc_direction_matrix,
         offset=0.45,
         width =0.15,
         color = "white",
         colnames_angle = 45,
         font.size = 2.5,
         colnames_offset_y = 0,
         colnames_offset_x = 0
         )+
  scale_fill_gradientn(colors=c("#053061", "#2166AC", "#FFFFFF", "#B2182B", "#530019"),
                      values= scales::rescale(c(-3.5e5, -1e5, 0, 1e5, 2.95e5)),
                      limits = c(-3.5e5, 2.95e5),
                      name = "Weighted Averaged \nLog-fold Change",
                      na.value = "grey78"
                      )+
  new_scale_fill()



ggsave( "data/Gen2R_tree_perOrder.pdf", t,
      device = cairo_pdf,
      width = 40, height = 30, units = "cm")

```

This tree and the information present in it is complicated, but it is the best I can do to represent such a complex dataset.
For us, this tree has been a great source of information and it takes some time to get into it and understand each information.
Once it's done though, a glance is enough to compare our results with results presented in many articles.


One of the reasons the ggtree package is great is because of the fantastic tutorials available on internet. I learned everything from [this page](https://yulab-smu.top/treedata-book/chapter7.html).

# 3 - Blast database creation and query

## Database Creation and query
--> See the script blast_db_creation_query.sh

## Get list of strains

```{r}
## Import output
alignment <- read.csv("save_me/ASV_isolates_alignment.csv")

## Clean up the results
strain_list <- alignment |>
  ## Remove short alignments
  filter(qstart < 30) |>
  ## Pull strain names only
  pull(stitle) |>
  ## remove duplicate strains
  unique()

## Print complete list of strain of interest
cat(sprintf("In the end, we have %d strains of interest to phenotype:\n",
            length(strain_list)), 
    strain_list)

```

The rest of the workflow included picking the strains of interest.
We did not have the means to screen that many strains, so we choosed based on the following critera:
- We pick **all the strains** that have a **perfect 16S/ASV match**,
- For each ASV that does not have a perfect match, we pick **1 strain with 1 mismatch**.
- We did **not** test the **filamentous bacteria** because they were unsuitable for our protocol, and the **16SrRNA is a mediocre identity marker for them**.

# End

If you want the whole story, the preprint and the link to **all the data and codes** are available on [BioRxiv](https://www.biorxiv.org/content/10.64898/2026.03.06.709792v1).


```{r session_info}
sessionInfo()

```
